Tools for identifying unexpectedly low microfossil count sums

### 2024-01-12 count_check_MS.qmd
gt

# tar_load("summarised_bird_singletons")
# tar_load("summarised_bird")
# tar_load("testate_summ")
 tar_load("last_chance_summ")
# tar_load("pollen_assem_res")
# tar_load("pollen_dataset_res")
 tar_load("pollen_results")
 tar_load("hershop_pollen_summ")
 tar_load("wildhorse")
 tar_load("pollen3")
 tar_load("pollen1")
 tar_load("pollen2")
# tar_load("owens_summ1")
# tar_load("owens_summ")
# tar_load("owens_est_summ")
 tar_load("diatom1_summ")
 tar_load("diatom2_summ")
 tar_load("chironomid1_summ")
 tar_load("marine1_summ")
# tar_load("chironomid2_summ")

@@

```{r}
#| label: tbl-multitaxa
#| tbl-cap: Properties of different data sets.
tar_read(count_summary) |> 
  #manually escape % as cannot use kbl(escape = TRUE) with linebreak
  mutate(across(where(is.character), str_replace, pattern = "%", replacement = "\\\\%")) |> 
  gt(x = _)
# , booktabs = TRUE, escape = FALSE,
#       col.names = linebreak(str_wrap(names(.), width = 10))
#     ) |> 
#   kable_styling(latex_options = c("basic", "hold_position"))
```

@@

Some of these datasets are older datasets which may, despite the available metadata, have been back-transformed from percent data or digitised with the loss of rare taxa and precision.
For example, all but `r hershop_pollen_summ$has_singleton |> word()` of the `r hershop_pollen_summ$n` assemblages in the archived pollen data from Hershop Bog, Texas have a greatest common divisor of two or more (the exception has a single singleton and is otherwise divisible by two). 
The original data [@Patty1968] were counted to exactly 200 pollen grains and fern spores. 
It seems likely that the data were converted to percent, rounded and then back-transformed to counts.
For most purposes, such back-transformed or digitised data are perfectly adequate, but the Chao 1 richness estimator calculated using the number of singletons and doubletons, would fail [@Chao1987].

>>>>

# tar_load("summarised_bird_singletons")
# tar_load("summarised_bird")
# tar_load("testate_summ")
 tar_load("last_chance_summ")
# tar_load("pollen_assem_res")
# tar_load("pollen_dataset_res")
 tar_load("pollen_results")
 tar_load("hershop_pollen")
 tar_load("wildhorse")
 tar_load("pollen3")
 tar_load("pollen1")
 tar_load("pollen2")
# tar_load("owens_summ1")
# tar_load("owens_summ")
# tar_load("owens_est_summ")
 tar_load("diatom1_summ")
 tar_load("diatom2_summ")
 tar_load("chironomid1_summ")
 tar_load("marine1_summ")
# tar_load("chironomid2_summ")

@@

```{r}
#| label: tbl-multitaxa
#| tbl-cap: Properties of different data sets.
tar_read(count_summary) |> 
  gt()
```

@@

Some of these datasets are older datasets which may, despite the available metadata, have been back-transformed from percent data or digitised with the loss of rare taxa and precision.
For example, all but `r hershop_pollen$has_singleton |> word()` of the `r hershop_pollen$n` assemblages in the archived pollen data from Hershop Bog, Texas have a greatest common divisor of two or more (the exception has a single singleton and is otherwise divisible by two). 
The original data [@Patty1968] were counted to exactly 200 pollen grains and fern spores. 
It seems likely that the data were converted to percent, rounded and then back-transformed to counts.
For most purposes, such back-transformed or digitised data are perfectly adequate, but the Chao 1 richness estimator [@Chao1987] calculated using the number of singletons and doubletons, would fail.
